The Unicode Cookbook \linebreak for Linguists

### 2017-12-27 book/chapters/ipa_meets_unicode.tex
deals with chapter 4 revisions -- wip

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel, which is encoded in IPA with a
`single-story' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
front unrounded vowel, which is encoded in IPA with a `double-story' <a>
\textsc{latin small letter a} at \uni{0061}, basically treating them as
homoglyphs, although they are different phonetic symbols. But even among
linguists this distinction leads to problems. For example, as pointed out by
\citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of
languages are reported to have a five-vowel system \citep{Maddieson1984}.
Historically, linguistic descriptions tend not to include precise audio
recording and measurements of formants, so this may lead one to ask if the many
<a> characters that are used in phonological description reflects a
transcriptional bias. The common use of <a> in transcriptions could be in part
due to the ease of typing the letter on an English keyboard (or for older
descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that even in many well-designed fonts the italics of <a> and
<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
the italics of these characters are highly similar <\textit{a}> and
<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
the only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}

@@

Finally, the IPA states that ``diacritics may be placed above a symbol with a
descender''. For example, for marking marking of voiceless pronunciation of
voiced segments the IPA uses the ring diacritic. Originally, the ring should be
placed below the base character, like in <m̥>, using the \textsc{combining ring
below} at \uni{0325}. However, in letters with long descenders the IPA also
allows to put the ring above the base, like in <ŋ̊>, using the \textsc{combining
ring above} at \uni{030A}. Yet, proper font design does not have any problem
with rendering the ring below the base character, like in <ŋ̥>, so for strict
IPA encoding we propose to standardize on the ring below. As a principle, for
strict IPA encoding only one options should be allowed for all diacritics.

@@

   \item Tone letters are normally written at the end of the syllable,
         possibly occurring immediately adjacent to a consonant in the coda of
         the syllable. Such tone markers should of course not be treated as
         belonging to this consonant, so we propose to treat tone letters as 
         separate segments.

@@

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization (see
Section~\ref{pitfall-canonical-equivalence}).
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at \uni{00E1}) will be split into its parts. This might
sound preferable for a linguistic analysis, as the different diacritics are
separated from the base characters. However, note that most attached elements
like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
<ç>) will be decomposed (see Section~\ref{pitfall-missing-decomposition}). In
general, Unicode decomposition does not behave like a feature decomposition as
expected from a linguistic perspective. It is thus important to consider Unicode
decomposition only as a technical procedure, and not assume that it is
linguistically sensible.

@@

By Unicode Canonical Combining Classes, the diacritics on top of a character,
like <\dia{0334}> (Combining Class number 1), always come before diacritics
below (Combining Class number 220), which in turn always come before diacritics
above (Combining Class number 230), which in turn come before diacritics over
multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
(Combining Class number 233). We follow this order, but add the other IPA
diacritics (which are not diacritics in the Unicode sense) between diacritics
below and the tie bar. Further, \textit{within} all these classes of diacritics
there is no canonical ordering specified by Unicode, so we propose an explicit
ordering here.

@@

With each revision of the IPA, many decisions need to be made by 
the Association as to which symbols should be added, removed or 
changed. For example, in the 1989 revision of the IPA at the Kiel Convention, 
changes to specific symbols (in previous charts) were debated and 
the Association's members certain decisions. The prevailing mood at 
the convention was not to change specific symbols unless a strong 
case was made \citep{Ladefoged1990a}. For example, two such decisions 
included:

\begin{itemize}
	\item symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
       because the latter were the symbols used by nearly all Khoisanists and
       Bantuists.
	\item the Americanist tradition of using using <\dia{030C}>, a
       \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
       in <š~ž~č~ǰ>, was not adopted because the Association members at the
       convention ``were not sufficiently impressed by arguments ... to the
       effect that these sounds formed a natural class, and thus is would be
       appropriate to recognize this by maintaining a common aspect to their
       symbolism'' \citep[62]{Ladefoged1990a}. 
\end{itemize}

\noindent These decisions have practical consequences for transcribers 
of IPA, particularly those who wish to follow recommended practices of 
encoding electronic text in the Unicode Standard. For example, the Unicode 
Standard contains removed symbols as labels them as clicks, e.g. <ʇ> 
\textsc{latin small letter turned t} at U+0287 in the \textsc{IPA Extensions} 
block has the comment `dental click (sound of ``tsk tsk'')'. In this case, 
the IPA transcriber must know the status of click symbols in the current 
version of the IPA and then identify those characters within the Unicode Standard.

@@

In the course of collecting a large sample of phoneme systems across the world's
languages \citep{Moran2012}, \citet{Moran_etal2014} found that in order 
to preserve distinctions both within and across language descriptions, 
additions to the approved IPA glyph set were needed. Wherever possible 
these additions were drawn from the extIPA symbols for disordered
speech \citep{Duckworth_etal1990}.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
This section describes these proposed additions to the IPA glyph set. The
additions are not part of the official IPA recommendations, so they should be 
used with caution.

>>>>

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel, which is encoded in IPA with a
``single-story'' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
front unrounded vowel, which is encoded in IPA with a ``double-story'' <a>
\textsc{latin small letter a} at \uni{0061}, basically treating them as
homoglyphs, although they are different phonetic symbols. But even among
linguists this distinction leads to problems. For example, as pointed out by
\citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of
languages are reported to have a five-vowel system \citep{Maddieson1984}.
Historically, linguistic descriptions tend not to include precise audio
recording and measurements of formants, so this may lead one to ask if the many
<a> characters that are used in phonological description reflects a
transcriptional bias. The common use of <a> in transcriptions could be in part
due to the ease of typing the letter on an English keyboard (or for older
descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that even in many well-designed fonts the italics of <a> and
<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
the italics of these characters are highly similar <\textit{a}> and
<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
the only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).} This 
approach was taken by the Language Science Press to distinguish 
between the two different orthographic <a>'s in Chakali in 
\cite{Brindle2017}.\footnote{\url{http://langsci-press.org/catalog/book/74}}

@@

Finally, the IPA states that ``diacritics may be placed above a symbol with a
descender''. For example, for marking marking of voiceless pronunciation of
voiced segments the IPA uses the ring diacritic. Originally, the ring should be
placed below the base character, like in <m̥>, using the \textsc{combining ring
below} at \uni{0325}. However, in letters with long descenders the IPA also
allows to put the ring above the base, like in <ŋ̊>, using the \textsc{combining
ring above} at \uni{030A}. Yet, proper font design does not have any problem
with rendering the ring below the base character, like in <ŋ̥>, so for strict
IPA encoding we propose to standardize on the ring below. As a principle, for
strict IPA encoding only one option should be allowed for all diacritics.

@@

   \item Tone letters <˥, ˦, ˧, ˨, ˩> from \uni{02E5}--\uni{02E9} 
   		 are normally written at the end of the syllable,
         possibly occurring immediately adjacent to a consonant in the coda of
         the syllable. Such tone markers should of course not be treated as
         belonging to this consonant, so we propose to treat tone letters as 
         separate segments.

@@

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization (see
Section~\ref{pitfall-canonical-equivalence}).
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at \uni{00E1}) will be split into its parts. This might
sound preferable for a linguistic analysis, as the different diacritics are
separated from the base characters. However, note that most attached elements
like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
(e.g.~in <ɝ>) will not be decomposed. In
general, Unicode decomposition does not behave like a feature decomposition as
expected from a linguistic perspective. It is thus important to consider Unicode
decomposition only as a technical procedure, and not assume that it is
linguistically sensible.

@@

By Unicode Canonical Combining Classes, overlay diacritics 
like <\dia{0334}> (Combining Class number 1), always come before diacritics
below (Combining Class number 220), which in turn always come before diacritics
above (Combining Class number 230), which in turn come before diacritics over
multiple characters like the tie bar <\dia{0361}{\large\fontspec{CharisSIL}◌}>
(Combining Class number 233). We follow this order, but add the other IPA
diacritics (which are not diacritics in the Unicode sense) between diacritics
below and the tie bar. Further, \textit{within} all these classes of diacritics
there is no canonical ordering specified by Unicode, so we propose an explicit
ordering here.

@@

With each revision of the IPA, many decisions need to be made by 
the Association as to which symbols should be added, removed or 
changed. For example, in the 1989 revision of the IPA at the Kiel Convention, 
changes to specific symbols (in previous charts) were debated and 
the Association's members made certain decisions. The prevailing mood at 
the convention was not to change specific symbols unless a strong 
case was made \citep{Ladefoged1990a}. For example, two such decisions 
included:

\begin{itemize}
	\item Symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
       because the latter were the symbols used by nearly all Khoisanists and
       Bantuists.
	\item The Americanist tradition of using using <\dia{030C}>, a
       \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
       in <š~ž~č~ǰ>, was not adopted because the Association members at the
       convention ``were not sufficiently impressed by arguments ... to the
       effect that these sounds formed a natural class, and thus is would be
       appropriate to recognize this by maintaining a common aspect to their
       symbolism'' \citep[62]{Ladefoged1990a}. 
\end{itemize}

\noindent These decisions have practical consequences for transcribers 
of IPA, particularly those who wish to follow current recommended practices of 
encoding electronic text in the Unicode Standard. For example, the Unicode 
Standard contains \textsc{latin small letter turned t} <ʇ> at U+0287, 
which is no longer part of the IPA. It still exists, however, in the Unicode 
\textsc{IPA Extensions} block, with the comment ``dental click (sound of `tsk tsk')''. 
In such cases, the IPA transcriber must know the status of legacy symbols in the current 
version of the IPA and the correct characters in the Unicode Standard.

@@

In the course of collecting a large sample of phoneme systems across the world's
languages, \citet{Moran_etal2014} found that in order 
to preserve distinctions both within and across language descriptions, 
additions to the approved IPA glyph set were needed. Wherever possible 
these additions were drawn from the extIPA symbols for disordered
speech \citep{Duckworth_etal1990}.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}
This section describes these proposed additions to the IPA glyph set. The
additions are not part of the official IPA recommendations, so they should be 
used with caution.
